---
title: "Lab 6: Unix Filters and Pipes"
subtitle: "Stat 133, Fall 2022"
output: 
  github_document:
    keep_html: true
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

> ### Learning Objectives:
>
> - Use Unix output redirection with `>` and `>>`
> - Use Unix pipes with `|`
> - Use filters: `cut`, `paste`, `sort`, `uniq`, `grep`

### General Instructions

- You DON'T need to use an `.Rmd` file for this lab
- Part of this lab involves writing code in an `R` script file.
- Name this file as `lab06-first-last.txt`, where `first` and `last` are your
first and last names (e.g. `lab06-gaston-sanchez.txt`).
- Submit your `.txt` file to bCourses, in the corresponding lab assignment.


__See cheatsheet at the end of this document__


-----


## Climbing Ropes Data

In this lab you are going to use data about __dynamic ropes__ better known
as "climbing ropes" (a few rows shown below)

```{r echo = FALSE, comment=""}
dat <- read.csv("ropes.csv")
head(dat)
```


The data comes in two file flavors:

- `ropes.txt` (space-separated values)

- `ropes.csv` (comma-separated values)


Use the climbing ropes data to perform the following tasks. All the 
commands have to be bash-shell commands (not R commands).

- `cut` allows you to _select_ columns
- `paste` allows you to _merge_ columns
- `sort` can be used to _arrange_ lines
- `sort` can also be used to _group by_ lines
- `sort` and `uniq` can be used to _count_ occurrences
- `grep` allows you to _filter_ rows


-----


## 1) Names of Ropes

__1.1)__ Here's a pipeline to obtain the names of the first 5 ropes in the 
`ropes.csv` data file. First we `cut` the first column (`Name`), then we skip 
the first row with `tail +2`, and finally we `head` the first 5 elements:

```{bash}
# display names of first 5 ropes
cut -f 1 -d "," ropes.csv | tail +2 | head -n 5
```

__1.2)__ We can use the previous pipeline, and redirect the output to a text 
file `names5.txt` using the output redirection operator `>`.

```{bash}
# file with names of first 5 ropes
cut -f 1 -d "," ropes.csv | tail +2 | head -n 5 > names5.txt
```

To confirm that `names5.txt` has the tent names, we can use `head` to display 
the first five lines of the created file:

```{bash}
head names5.txt
```

__1.3)__ We can take the pipeline above, and pipe it with `sort` so that we 
get the names sorted alphabetically

```{bash}
# display names of first 5 ropes, sorted alphabetically
cut -f 1 -d "," ropes.csv | tail +2 | head -n 5 | sort
```


### 1.1) Your turn:

__1.4)__ Modify the previous pipe to get the names of the first 5 ropes,
from the file `ropes.txt`, sorting values alphabetically in decreasing order

```{bash}
# your bash command
```


__1.5)__ Using `ropes.txt`, write a pipe to get the names of the last 5 ropes 
(no sorting required)

```{bash}
# your bash command
```


__1.6)__ Using `ropes.txt`, write a pipe to get the `Name` and `Brand` of the 
first 5 ropes

```{bash}
# your bash command
```


__1.7)__ Write a pipe to get the `Name`, `Diameter`, and `Weight` of the last 5 
ropes in `ropes.txt`

```{bash}
# your bash command
```


-----


## 2) Listing and Counting Categories

__2.1)__ Often, we want to list unique values. For example, say we are interested 
in knowing which are the (unique) categories of `Brand`. We can pipe `sort` and 
`uniq` for this purpose. Here's how:

```{bash}
cut -f 2 -d "," ropes.csv | tail +2 | sort | uniq
```

__2.2)__ What if, in addition to identify the unique categories, we want to 
obtain their frequencies (i.e. counts)? Then, simply add the option `-c` for `uniq`

```{bash}
cut -f 2 -d "," ropes.csv | tail +2 | sort | uniq -c
```


### 2.1) Your turn:

__2.3)__ Take the previous command and pipe it again with `sort` in order to 
display the `Brand` categories (and their counts) in increasing order

```{bash}
# your bash command
```


__2.4)__ With `ropes.csv`, write a pipe to list (i.e. display) the unique 
categories of `Falls`

```{bash}
# your bash command
```


__2.5)__ With `ropes.csv`, write a pipe to list (i.e. display) the unique 
categories of `Falls` and their counts

```{bash}
# your bash command
```


-----


## 3) Selecting Rows

__3.1)__ Another useful unix filter is `grep`. One basic use for this command is 
to match certain pattern. For example, here's how to find (match) those rows 
of ropes with brand `Beal`.

```bash
grep "Beal" ropes.csv
```

We can pipe the previous command with `wc` to count how many `Beal` ropes are 
in `ropes.csv` (i.e. counting number of lines):

```{bash}
grep "Beal" ropes.csv | wc -l
```


__3.2)__ If you want to create a file with the data for `Beal` ropes, you can 
redirect the output to a text file `beal-ropes.csv`:

```bash
grep "Beal" ropes.csv > beal-ropes.csv
```

Notice that the created file `beal-ropes.csv` does NOT contain names of columns 
in the first row. If you want the file to include column names, you need to 
write two commands like so:

```bash
head -n 1 ropes.csv > beal-ropes.csv
grep "Beal" ropes.csv >> beal-ropes.csv
```


### 3.1) Your Turn

__3.3)__ With `ropes.txt`, pipe `grep` with `wc` to count the number of ropes 
from brand `Edelrid`

```{bash}
# your bash command
```


__3.4)__ With `ropes.txt`, write a similar pipe to the one above to find how 
many `Sterling` ropes are in this file

```{bash}
# your bash command
```


__3.5)__ With `ropes.txt`, write a pipe to create a file `sterling-ropes.txt` 
containing the data for ropes of brand `Sterling`:

```{bash}
# your bash command
```


__3.6)__ With `ropes.txt`, write commands to create a file `pmi-ropes.txt` 
containing the data for ropes of brand `PMI`, arranged by name alphabetically. 
This file should have column names in the first line (i.e. first row).

```{bash}
# your bash command
```


-----


## 4) More Pipes (your turn)

__4.1)__ With `ropes.txt`, write a pipeline to obtain the unique categories of 
`Brand` in decreasing order. The output of the pipeline should contain only the 
categories (NOT the counts). _HINT:_ `cut`, `sort`, and `uniq` are your friends.

```{bash}
# your bash command
```


__4.2)__ With `ropes.csv`, write a pipeline to obtain the counts (i.e. 
frequencies) of the different `Weight` values (column 4), displayed from 
largest to smallest (i.e. descending order). The first column corresponds to 
count, the second column corresponds to weight. Redirect the output to a text 
file `weight-counts.txt`. _HINT:_  `cut`, `tail`, `sort`, `uniq`; and 
redirection `>` operator, are your friends.

```{bash}
# your bash command
```


__4.3)__ With `ropes.csv`, use output redirection commands to create a CSV file 
`mammut-ropes.csv` containing data for the `Mammut` brand. Your CSV file should 
include column names in the first row. 
_HINT:_ redirection operators `>` and `>>`, as well as `head` and `grep` are your friends.

```{bash}
# your bash command
```


__4.4)__ Use the previously created file `mammut-ropes.csv` to select, 
in separate commands, the columns `Name`, `Weight`, and `Force`. Store each 
column in its own text file: `name.txt`, `weight.txt`, and `force.txt`

```{bash}
# your bash command
```


__4.5)__ Use the previously created files `name.txt`, `weight.txt`, and 
`force.txt` to `paste` them (i.e. merge them), in that order, into a new CSV 
file `mammut-forces.csv` (comma separated values!!!).

```{bash}
# your bash command
```


__4.6)__ Use `ropes.txt`, to write a pipeline to list (display) the five 
heaviest ropes (their Name, Brand, and Weight) in decreasing order (by weight).

```{bash}
# your bash command
```


-----


## Cheat sheet: Unix filters


| Command                     | Description                   | R alternative             |
|-----------------------------|-------------------------------|---------------------------|
| `wc nba2017-players.csv`    | count lines, words, and bytes | `object.size()`, `nrow()` |
| `wc -l nba2017-players.csv` | count number of lines         | `nrow()`                  |
| `head nba2017-players.csv`  | inspect first 10 rows         | `head()`                  |
| `tail nba2017-players.csv`  | inspect last 10 rows          | `tail()`                  |
| `less nba2017-players.csv`  | see conropes with a paginator | `View()`                  |


Extracting columns with __`cut`__

| Option    | Description                                        |
|-----------|----------------------------------------------------|
|  `-f` 1,5 | return columns 1 and 5, delimited by tabs.         |
|  `-f` 1,5 | return columns 1 through 5, delimited by tabs.     |
|  `-d ","` | use commas as the delimiters.                      |
|  `-c 2-7` | return characters 2 through 7 from the file.       |


Sorting lines with __`sort`__

| Option | Description                                          |
|--------|------------------------------------------------------|
|  `-n`  | sort in numerical order rather than alphabetically.  |
|  `-r`  | sort in reverse order, z to a or decreasing numbers. |
|  `-f`  | fold uppercase into lowercase (i.e. ignore case).    |
|  `-u`  | return a unique representative of repeated items.    |
|  `-k 3`| sort lines based on column 3 (tab or space delimiters) |
|  `-t ","` | use commas for delimiters.                      |
|  `-b`  | ignore leading blanks.                               |
|  `-d`  | sort in dictionary order.                            |


Isolating unique lines with __`uniq`__

| Option  | Description                                        |
|---------|----------------------------------------------------|
|  `-c`   | adds a count of how many times each line occurred. |
|  `-u`   | lists only lines that are not repeated.            |
|  `-d`   | lists only lines that are duplicated.              |
|  `-i`   | ignore case when determining uniqueness            |
|  `-f 4` | ignore the first 4 fields (space delimiter)        |
