---
title: 'Lab 2: Intro to R Markdown `Rmd` Files'
subtitle: Stat 133
output:
  html_document:
    toc: yes
  pdf_document:
    toc: yes
---

> ### Learning Objectives:
>
> - Get started with `.Rmd` files
> - Get to know markdown syntax

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



## 1) Introduction to R Markdown files

Most of the times you won't be working directly on the console.
Instead, you will be typing your commands in some __source file__.
The most basic type of source files are known as _R script files_. 
But there are more flavors of source files. A very convenient type of source
file that allow you to mix R code with narrative is an __R markdown file__
commonly referred to as `Rmd` file.


### 1.1) Get to know the `Rmd` files

In the menu bar of RStudio, click on __File__, then __New File__,
and choose __R Markdown__. Select the default option (Document),
and click __Ok__.

__Rmd__ files are a special type of file, referred to as a _dynamic document_,
that allows to combine narrative (text) with R code. Because you will
be turning in most homework assignments as `Rmd` files, it is important
that you quickly become familiar with this resource.

Locate the button __Knit HTML__ (the one with a knitting icon) and click on it
so you can see how `Rmd` files are renderer and displayed as HTML documents.

_R markdown_ files use a special syntax called __markdown__. To be more precise, 
Rmd files let you type text using either: 1) R syntax for code that needs to be 
executed; 2) markdown syntax to write your _narrative_, and 3) latex syntax for 
math equations and symbols.

You will have time to learn the basics of this syntax during the semester, and
we expect that feel comfortable with markdown at the end of the course.


-----


## 2) Markdown Syntax Tutorial

RStudio has a very comprehensive R Markdown tutorial:
http://rmarkdown.rstudio.com/

Your GSI will go over the basics of Rmd files, which involve the following sections:

- Introduction: https://rmarkdown.rstudio.com/lesson-1.html

- How it Works: https://rmarkdown.rstudio.com/lesson-2.html

- Code Chunks: https://rmarkdown.rstudio.com/lesson-3.html

- Inline Code: https://rmarkdown.rstudio.com/lesson-4.html

- Markdown Basics: https://rmarkdown.rstudio.com/lesson-8.html

 -R Markdown cheatsheet (also available in bCourses; see `Files/cheatsheets/`): https://github.com/rstudio/cheatsheets/raw/main/rmarkdown-2.0.pdf


-----


## 3) Your turn: Pythagoras formula

Open a new `Rmd` file in the _source_ pane, and use it to compute the following.

The pythagoras formula is used to compute the length of the hypotenuse, $c$, of 
a right triangle with legs of length $a$ and $b$.

$$
c = \sqrt{a^2 + b^2}
$$

Calculate the hypotenuse of a right triangle with legs of length 3 and 4.
Use the `sqrt()` function, and create variables `a = 3` and `b = 4`.
If you don't know what's the symbol to calculate exponents, search for the 
help documentation of the arithmetic operators: `?Arithmetic`.

```{r}
a <- 3
b <- 4
sqrt(a^2 + b^2)

```


-----


## 4) Binomial Formula

The formula for the binomial probability is:

$$
Pr(k; n, p) = Pr(X = k) = {n \choose k} p^k (1-p)^{n-k}
$$

where:

- $n$ is the number of (fixed) trials
- $p$ is the probability of success on each trial
- $1 - p$ is the probability of failure on each trial
- $k$ is a variable that represents the number of successes out of $n$ trials
- the first term in parenthesis is not a fraction, it is the number of 
combinations in which $k$ success can occur in $n$ trials

R provides the `choose()` function to compute the number of combinations: 

$$
{n \choose k} = \frac{n (n-1) \cdots (n - k +1)}{k (k-1) \cdots 1}
$$

For instance, the number of combinations in which $k$ = 2 
success can occur in $n$ = 5 trials is:

```{r}
choose(n = 5, k = 2)
```

Combinations are typically expressed in terms of factorials as:

$$
\frac{n!}{k! (n - k)!}
$$

Conveniently, R also provides the function `factorial()` to calculate the 
factorial of an integer:

```{r}
factorial(4)
```



### 4.1) Your Turn: Binomial Formula

Let's consider a simple example. A fair coin is tossed 5 times. What is the 
probability of getting exactly 2 heads?

a) Create the objects `n`, `k`, and `p` for the number of trials, the number of 
success, and the probability of success, respectively.

```{r}
n <- 5
k <- 2
p <- 0.5

```


b) Use `factorial()` to compute the number of combinations "_n choose k_"

```{r}
factorial(n)/(factorial(k)*factorial(n-k))

```


c) Apply the binomial formula, using `factorial()`, to calculate the probability 
of getting exactly 2 heads out of 5 tosses.

```{r}
(factorial(n)/(factorial(k)*factorial(n-k))) * (p^k) * (1-p)^(n-k)

```


d) Recalculate the same probability but now using `choose()` 
(instead of `factorial()`)

```{r}
choose(n,k) * (p^k)*(1-p)^(n-k)

```


e) Consider rolling a fair die 10 times. What is the probability 
of getting exactly 3 sixes?

```{r}
choose(n=10,k=3) * (1/6)^3 * (5/6)^7

```


f) Now look for help documentation (e.g. `help.search()` or `??`) using the 
keyword binomial: `binomial`.
You should get a list of topics related with the searched term `binomial`.
Choose the one related with the _Binomial Distribution_, which is 
part of the R package `stats` (i.e. `stats::Binomial`).

Read the documentation and figure out how to use the `dbinom()` function 
to obtain the above probabilities: 2 heads in 5 coin tosses, and 
3 sixes in 3 rolls of a die.

```{r}
dbinom(2, 5, 0.5)
dbinom(3, 3, 1/6)

```


j) How would you modify the previous binomial function to calculate the 
same probability (2 heads in 5 tosses) of a __biased__ coin with a chance of 
heads of 35%?

```{r}
dbinom(2, 5, .35)

```


k) Finally, obtain the probability of getting more than 3 heads in 5 tosses
with a biased coin of 35% chance of heads.

```{r}
sum(dbinom(3:5,5,.35))
# or alternatively
pbinom(5,5,.35) - pbinom(2,5,.35)

```


-----


## 5) Getting Help

Because we work with functions all the time, it's important to know certain
details about how to use them, what input(s) is required, and what is the
returned output.

There are several ways to get help.

If you know the name of a function you are interested in knowing more,
you can use the function `help()` and pass it the name of the function you
are looking for:

```r
# documentation about the 'abs' function
help(abs)

# documentation about the 'mean' function
help(mean)
```

Alternatively, you can use a shortcut using the question mark `?` followed
by the name of the function:

```r
# documentation about the 'abs' function
?abs

# documentation about the 'mean' function
?mean
```

- How to read the manual documentation
	- Title
	- Description
	- Usage of function
	- Arguments
	- Details
	- See Also
	- Examples!!!

`help()` only works if you know the name of the function your are looking for.
Sometimes, however, you don't know the name but you may know some keywords.
To look for related functions associated to a keyword, use double
`help.search()` or simply `??`

```r
# search for 'absolute'
help.search("absolute")

# alternatively you can also search like this:
??absolute
```
Notice the use of quotes surrounding the input name inside `help.search()`




### 5.1) Your turn

Type commands directly on the console:

a) There are several tabs in the pane `Files, Plots, Packages, Help, Viewer`.
Find what does the tab __Files__ is good for?

* The __Files__ tab is useful for navigating the files on your computer (like Windows Explorer) in order to find the files required for your project.

b) What about the tab __Help__? 

* The __Help__ tab is useful for finding out exactly how some built-in R functions work, what their arguments are, and even how to find functions that you might not know the exact name of.

c) In the tab __Help__, what happens when you click the button with a House icon?

* It takes you to the home page of the help section with resources and manuals.

d) Now go to the tab __History__. What is it good for? and what about the 
buttons of its associated menu bar?

* The __History__ tab displays all of the commands you have entered in the console in chronological order.

e) Likewise, what can you say about the tab __Environment__?

* The __Environment__ tab displays the objects (their names and their corresponding values) that are currently in the environment.



