---
title: 'Lab 1: Getting started with R and RStudio'
subtitle: "Stat 133"
output: 
  html_document:
    toc: true
---

> ### Learning Objectives:
>
> - Get started with R as a scientific calculator
> - Understand pane layout of RStudio
> - Understand the help documentation in R
> - How to install packages

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



## 1) R and RStudio

- You should have installed __R__
	- R for Mac: [https://cran.r-project.org/bin/macosx/](https://cran.r-project.org/bin/macosx/)
	- R for windows: [https://cran.r-project.org/bin/windows/base/](https://cran.r-project.org/bin/windows/base/)

- You should also have installed __RStudio__
	- RStudio download: [https://www.rstudio.com/products/rstudio/download/](https://www.rstudio.com/products/rstudio/download/)

- R and RStudio are two different things
	- R is both a software and language (like python, matlab, julia, etc)
	- R comes with its minimal Graphical User Interface (R-GUI), but we won't use this.
	- Instead, we are going to use RStudio as our main working environment. 
	- RStudio is much more than a simple GUI; RStudio is technically an _Integrated Development Environment_ (IDE).
	- We are going to use mainly RStudio, which provides a nice working environment and development framework.



## 2) R as a scientific calculator

Launch RStudio and notice the default position of the panes (or panels):

- Console (entire left)
- Environment/History (tabbed in upper right)
- Files/Plots/Packages/Help (tabbed in lower right)

__FYI:__ you can change the default location of the panes, 
among many other things: [Customizing RStudio](https://support.rstudio.com/hc/en-us/articles/200549016-Customizing-RStudio). If you have no experience working with R/RStudio, you don't have to customize anything right now. It's better if you wait some days until you get a better feeling of the working environment. You will probably be experimenting (trial and error) some time with the customizing options until you find what works for you.


### 2.1) First contact with the R console

If you have never used software in which you have to type commands and code, our best suggestion is that you begin typing basic things in the __console__, using R as a scientific calculator.

For instance, consider the monthly bills of Leia (a fictitious stats undergrad student): 

- cell phone $80
- transportation $20
- groceries $527
- gym $10
- rent $1500
- other $83

You can use R to find Leia's total expenses by typing these commands in the console:

```{r}
# total expenses
80 + 20 + 527 + 10 + 1500 + 83
```

Often, it will be more convenient to create __objects__ or __variables__ 
that store one or more values. To do this, type the name of the variable, 
followed by the assignment operator `<-`, followed by the assigned value.
For example, you can create an object `phone` for the cell phone bill, 
and then inspect the object by typing its name:

```{r}
# objects with assigned values
phone <- 80
phone
```

All R statements where you create objects are known as "assignments", and they have this form:

```r
object <- value
```

this means you assign a `value` to a given `object`;
you can read the previous assignment as "phone gets 80".

RStudio has a keyboard shortcut for the arrow operator `<-`: 
`Alt` + `-` (the minus sign).

Notice that RStudio automagically surrounds `<-` with spaces, 
which demonstrates a useful code formatting practice. So do yourself 
(and others) a favor by ALWAYS surrounding an assignment operator with spaces.

Alternatively, you can also use the equals sign `=` as an assignment 
operator:

```{r}
coffee = 30
coffee
```

You will be working with RStudio a lot, and you will have time to learn
most of the bells and whistles RStudio provides. Think about RStudio as
your "workbench". Keep in mind that RStudio is NOT R. RStudio is an environment
that makes it easier to work with R, while taking care of many of
the little tasks than can be a hassle.



### 2.2) Your Turn

a) Make more assignments to create variables `transportation`, `groceries`, `gym`, 
`rent`, and `other` with their corresponding amounts.

```{r}
# create variables
transportation <- 10
groceries <- 280
gym <- 65
rent <- 1500
```

b) Now that you have all the variables, create a `total` object with the 
sum of the expenses.

```{r}
# total expenses
total <- transportation + groceries + gym + rent
```

c) Assuming that Leia has the same expenses every month, how much would she 
spend during a school "semester"? (assume the semester involves five months).

```{r}
# semester expenses
semester <- total * 5
```

d) Maintaining the same assumption about the monthly expenses, how much would Leia
spend during a school "year"? (assume the academic year is 10 months).

```{r}
# year expenses
year <- semester * 2
```



### 2.3) Object Names

There are certain rules you have to follow when creating objects and variables.
Object names cannot start with a digit and cannot contain certain other characters 
such as a comma or a space. You will be wise to adopt a convention for demarcating 
words in names.


```r
i_use_snake_case
other.people.use.periods
evenOthersUseCamelCase
```

The following are invalid names (and invalid assignments)

```
# cannot start with a number
5variable <- 5

# cannot start with an underscore
_invalid <- 10

# cannot contain comma
my,variable <- 3

# cannot contain spaces
my variable <- 1
```

This is fine but a little bit too much:

```r
this_is_a_really_long_name <- 3.5
```


### 2.4) Functions

R has many functions. To use a function type its name followed by parenthesis.
Inside the parenthesis you pass an input. Most functions will produce some
type of output:

```r
# absolute value
abs(10)
abs(-4)

# square root
sqrt(9)

# natural logarithm
log(2)
```


### 2.5) Comments in R

All programming languages use a set of characters to indicate that a
specifc part or lines of code are __comments__, that is, things that are
not to be executed. R uses the hash or pound symbol `#` to specify comments.
Any code to the right of `#` will not be executed by R.

```r
# this is a comment
# this is another comment
2 * 9

4 + 5  # you can place comments like this
```


### 2.6) Case Sensitive

R is case sensitive. This means that `phone` is not the same
as `Phone` or `PHONE`

```{r}
# case sensitive
phone <- 80
Phone <- -80
PHONE <- 8000

phone + Phone
PHONE - phone
```


### 2.7) Your turn

a) Take your objects (i.e. variables) `phone`, `transportation`, `groceries`, 
`gym`, `rent`, and `other` and pass them inside the _combine_ function `c()` 
to create a vector `expenses`.

```{r}
# vector expenses
expenses <- c(phone, transportation, groceries, gym, rent)
M <- c("phone", "transportation", "groceries", "gym", "rent")
```


b) Now, use the graphing function `barplot()` to produce a barchart of 
`expenses`:

```{r}
# barchart
barplot(expenses)
```


c) You can see the manual (help) document of `barplot()` by typing `?barplot`;
look at the arguments of this function and learn how to use `col` to color the bars

```{r}
# barchart with colored bars
barplot(height = expenses, col = "black")
```


d) Use the argument `horiz` to get a barchart in horizontal orientation

```{r}
# barchart horizontally oriented
barplot(height = expenses, horiz = TRUE)
```


e) Use `names.arg` to add labels to the bars (with names of the variables below 
each of the bars)

```{r}
# barchart with bar labels
barplot(height = expenses, names.arg = M)
```


f) Find out how to use `sort()` to sort the elements in `expenses`, in order to
produce a bar-chart with bars in decreasing order. Also 
optional, see if you can find out how to display the values of each variable
at the top of each bar.

```{r}
# barchart in decreasing/increasing order
barplot(sort(expenses))
barplot(height = expenses, names.arg = M, col = "green", border = "black", xlab = "Category", ylab = "Cost ($)")
```


-----


## 3) Installing Packages

R comes with a large set of functions and packages. A package is a collection
of functions that have been designed for a specific purpose. One of the great
advantages of R is that many analysts, scientists, programmers, and users
can create their own packages and make them available for everybody to use them.
R packages can be shared in different ways. The most common way to share a
package is to submit it to what is known as __CRAN__, the
_Comprehensive R Archive Network_.

You can install a package using the `install.packages()` function. To do this, 
we recommend that you run this command directly on the console. Do NOT include 
this command in a code chunk of an Rmd file: you will very likely get an error
message when knitting the Rmd file.

To use `install.packages()` just give it the name of a package, surrounded by 
quotes, and R will look for it in CRAN, and if it finds it, R will download it 
to your computer.

```r
# installing (run this on the console!)
install.packages("knitr")
```

You can also install a bunch of packages at once:

```r
# run this command on the console!
install.packages(c("readr", "ggplot2"))
```

Once you installed a package, you can start using its functions by _loading_ 
the package with the function `library()`. By the way, when working on an Rmd 
file that uses functions from a given package, you MUST include a code chunk 
with the `library()` command.

```r
# (this command can be included in an Rmd file)
library(knitr)
```


### 3.1) Your turn

Type commands directly on the console:

a) On the console, install packages `"stringr"`, `"RColorBrewer"`, and "`xml2`"

b) Calculate: $3x^2 + 4x + 8$ when $x = 2$

c) Calculate: $3x^2 + 4x + 8$ but now with a numeric sequence for $x$ 
using `x <- -3:3`

