---
title: "Lab 7: Basic Data Objects"
subtitle: "Stat 133, Spring 2023"
output: 
  html_document:
    toc: true
---

> ### Learning Objectives
>
> - Work with vectors of different data types
> - Understand the concept of _atomic_ structures
> - Learn how to subset and slice R vectors
> - Understand the concept of _vectorization_
> - Understand _recycling_ rules in R

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE)
```


### General Instructions

- Write your descriptions, explanations, and code in an `Rmd` (R markdown) file.
- Name this file as `lab07-first-last.Rmd`, where `first` and `last` are your
first and last names (e.g. `lab07-gaston-sanchez.Rmd`).
- Knit your `Rmd` file as an html document (default option).
- Submit your `Rmd` and `html` files to bCourses, in the corresponding lab assignment.



## 1) Data: US State Facts and Figures (1970s)

Some weeks ago we looked at a couple of approaches to manipulate tables and create 
statistical graphics via `"dplyr"` and `"ggplot2"`. As you now know, the default 
object that R provides to handle data tables is the so called `data.frame`. 
It turns out that data frames are formed by a set of R vectors, which is the 
main topic of today's session.

In this lab, you are going to work with the data set `state` that has to with
US State Facts and Figures in the 1970s. Specifically, we are going to use the 
following objects:


```{r}
# vectors
name = state.name
area = state.area
region = state.region
population = state.x77[ ,"Population"]
income = state.x77[ ,"Income"]
illiteracy = state.x77[ ,"Illiteracy"]
```


### 1.1) Your turn: data-types

Use the function `typeof()` to see the data type of each of the above vectors.

```{r}
typeof(name)
typeof(area)
typeof(region)
typeof(population)
typeof(income)
typeof(illiteracy)

```


-----


## 2) Subsetting

Create a vector `four` by selecting the first four elements in `name`:

```{r}
four <- head(name, n = 4)
```

Single brackets `[ ]` are used to subset (i.e. subscript, split, slice) vectors.
For instance, to extract the first element in `name` we pass `1` inside the
brackets:

```{r}
four[1]
```


### 2.1) Your turn: subsetting vectors

Without running the code, try to guess the output of the following commands, and 
then run them to check your guess:

a) number 2: `four[2]`

```{r}
# Guess: "Alaska"
four[2]
```


b) an index of zero: `four[0]`?

```{r}
# Guess: NA
four[0]
# Answer: character(0)
```


c) a negative index: `four[-1]`?

```{r}
# Guess: "Arkansas"
four[-1]
# Answer: "Alaska"   "Arizona"  "Arkansas"
# or everything but the first element
```


d) various negative indices: `four[-c(1,2,3)]`?

```{r}
# Guess: "Arkansas"
four[-c(1,2,3)]
```


e) an index greater than the length of the vector: `four[5]`?

```{r}
# Guess: NA
four[5]
```


f) repeated indices: `four[c(1,2,2,3,3,3)]`?

```{r}
# Guess: "Alabama" "Alaska" "Alaska" "Arizona" "Arizona" "Arizona"
four[c(1,2,2,3,3,3)]
```


-----


## 3) Sequences

Often, you will need to generate vectors of numeric sequences, like the first 
five elements `1:5`, or from the first till the last element `1:length(name)`. 
R provides the colon operator `:`, and the functions `seq()`, and `rep()` 
to create various types of sequences.


### 3.1) Your turn: sequences and repetitions

Figure out how to use `seq()`, and bracket notation, to extract:

a) all the even elements in `name` (i.e. extract positions 2, 4, 6, etc)

```{r}
name[seq(2,length(name),2)]

```


b) all the odd elements in `area` (i.e. extract positions 1, 3, 5, etc)

```{r}
area[seq(1,length(area),2)]

```


c) all multiples of 5 (e.g. 5, 10, 15, etc) of `income`

```{r}
income[seq(5,length(income),5)]

```


d) elements in positions 10, 20, 30, 40, etc of `population`

```{r}
population[seq(10,length(population),10)]

```


e) all the even elements in `name` but this time in reverse order;
_hint_ the `rev()` function is your friend.

```{r}
rev(name[seq(2,length(name),2)])

```


-----


## 4) Logical Subsetting and Comparisons

Another kind of subsetting/subscripting style is the so-called 
__logical subsetting__. This kind of subsetting typically takes place when 
making comparisons. 

A __comparison operation__ occurs when you use comparison operators such as: 

- `>` greater than
- `>=` greater than or equal
- `<` less than
- `<=` less than or equal
- `==` equal
- `!=` different

For example:

```{r}
area_four <- area[1:4]

# elements greater than 100000
area_four[area_four > 100000]

# elements less than 100000
area_four[area_four < 100000]

# elements less than or equal to 53000
area_four[area_four <= 53000]

# elements different from 50000
area_four[area_four != 50000]
```

In addition to using comparison operators, you can also use __logical operators__ 
to produce a logical vector. The most common type of logical operators are:

- `&` AND
- `|` OR
- `!` negation

Run the following commands to see what R does:

```{r}
# AND
TRUE & TRUE
TRUE & FALSE
FALSE & FALSE

# OR
TRUE | TRUE
TRUE | FALSE
FALSE | FALSE

# NOT
!TRUE
!FALSE
```

Logical operators allow you to combine several comparisons:

```{r}
# vectors for first 10 elements
name10 <- name[1:10]
area10 <- area[1:10]
pop10 <- population[1:10]
inc10 <- income[1:10]

# names of first 10 states with area greater than 100,000
name10[area10 > 100000]

# names of first 10 states with populations between 2000 and 3000 (exclusive)
name10[pop10 > 2000 & pop10 < 3000]
```


### 4.1) Your turn: logical subsetting

Write commands, using bracket notation, to answer the following questions 
(you may need to use `is.na()`, `min()`, `max()`, `which()`, `which.min()`, `which.max()`):

a) name of states from region `"West"`

```{r}
name[region == "West"]

```


b) name of states from regions `"West"` or `"Northeast"`

```{r}
name[region == "West" | region == "Northeast"]

```


c) number (count) of states in the `"South"` region; _hint_: the `sum()`
function is your friend

```{r}
sum(!is.na(region[which(region == "South")]))

```


d) name of state with largest area; _hint_: the `which.max()`
function is your friend

```{r}
name[which.max(area)]

```


e) largest Population of states in the `"West"` region

```{r}
max(population[which(region == "West")])

```


f) name of states with illiteracy percentage of at most 60%, and population of 
at least 2000 

```{r}
name[illiteracy <= 0.6 & population >= 2000]

```


g) name of state in the `"Northeast"` with income equal to the median income 
of states in the `"Northeast"`

```{r}
name[income == median(income[region == "Northeast"])]

```


-----


## 5) Matrices

A vector can be extended into a 2-dimensional object in the form of a __matrix__ 
(which is a 2-dimensional array).

For instance, let's create a matrix with the `income` and `illiteracy` values 
of the first five states:

```{r}
II <- matrix(c(income[1:5], illiteracy[1:5]), nrow = 5, ncol = 2)
II
```

You can give names to rows and columns with the functions `rownames()` and `colnames()`:

```{r}
rownames(II) <- name[1:5]
colnames(II) <- c("income", "illiteracy")
II
```


### 5.1) Your Turn

a) Use the column-binding function `cbind()` to create a matrix `AP` with 
columns `area` and `population`, for the first 10 states; display `AP`

```{r}
AP <- cbind(area, population)
AP[1:10,]
```


b) Use the function `matrix()` to create a matrix `west` with `income` and
`illiteracy` values for states in the `"West"`. You will need to provide an
input vector combining `income` and `illiteracy` of the appropriate states.
Give meaningful row and column names.

```{r}
west <- matrix(cbind(income[region=="West"], illiteracy[region=="West"]), ncol = 2)
rownames(west) <- name[region=="West"]
colnames(west) <- c("income", "illiteracy")
west

```


c) If you create a matrix `NAR` by column-binding `name`, `area` and `region` 
for all states, what is the resulting data type---returned by `typeof()`---of
`NAR`?

```{r}
NAR <- matrix(cbind(name, area, region), ncol = 3)
typeof(NAR)
```


